---
title: "MANOVA"
output: html_document
date: "2025-11-16"
---

This is the evaluation of the experiment's results. The focus of this notebook is on answering the research question by conducting a multivariate analysis of variance (MANOVA) on the foundation of the measurement model which has been validated previously using an ordinal confirmatory factor analysis (CFA).

This notebook expands the master thesis by data analysis and respective code as well as short explanations. This markdown does not replace the thesis, it serves solely as an appendix.

<h1>Library Downloads</h1>
```{r}
library(tidyverse)
library(broom)
library(ggplot2)
library(dplyr)
library(finalfit)
library(mice)
library(miceadds)
library(MASS)
library(lavaan)
library(lavaan.mi)
library(semptools)
library(semTools)
library(psych)
library(corrplot)
```

<h1>Sample Load</h1>
First, the previously imputed data is loaded as a mids object:

```{r}
getwd()
final_imputed_data <- read_rds("data/mi_mids_object.rds")
glimpse(final_imputed_data)
```






